\textbf{Automated Reasoning\\Practical Assignment -- Part 2}

### 2022-12-15 part2/report/paper.tex
Add mpo docs

\usepackage{amsmath}
\usepackage{fancyhdr}
\usepackage{minted}
\usepackage{enumitem}
\usepackage{hyperref}
\usepackage{multirow}
\usepackage[table]{xcolor}
\usepackage{svg}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{scalerel}
\usepackage{amsfonts}

@@

\section{Robot} % (fold)
\label{sec:robot}

@@

	On grid (a) our program finds that the minimal solution uses 14 steps. The assignment of directions is visualized in figure \ref{robot-grid-a-solution}

	On grid (b) our program also finds a minimal solution of 14 steps, which is visualised in figure \ref{robot-grid-b-solution}

	On grid (c) our program also finds a minimal solution of 18 steps, which is visualised in figure \ref{robot-grid-c-solution}

@@

% section robot (end)

>>>>

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{fancyhdr}
\usepackage{minted}
\usepackage{enumitem}
\usepackage{hyperref}
\usepackage{multirow}
\usepackage[table]{xcolor}
\usepackage{svg}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{scalerel}
\usepackage{amsfonts}

@@

\pagebreak
\section{Robot} % (fold)
\label{sec:robot}

@@

	On grid (a) our program finds that the minimal solution uses 14 steps. The
	assignment of directions is visualized in figure
	\ref{robot-grid-a-solution}. On grid (b) our program also finds a minimal
	solution of 14 steps, which is visualised in figure
	\ref{robot-grid-b-solution}. On grid (c) our program also finds a minimal
	solution of 18 steps, which is visualised in figure
	\ref{robot-grid-c-solution}.

@@

\pagebreak
\section{Multiset Path Ordering}
\subsection{Explanation}
In this problem we are given a set of inequalities. These inequalities contain
terms which are either variables or functions over other terms. Our task is to
find a Multiset Path Ordering (MPO), which is an ordering on terms generated
from a total quasi-ordering relation $\trianglerighteq$ on the function
symbols, using multiset comparison.

To make our lives a bit easier we created a parser which can parse the
inequalities. To this end we created four data structures which can represent:
terms, functions, variables and inequalities. Notably, every one of these data
structures has a function \texttt{subterms} which recursively goes down the
term and gives all of its decendents. This is useful for when we define the
relations on terms.

We define the relations $\trianglerighteq, \rhd, \equiv$ on functions by
creating a new boolean variable for every relation and every pair of functions,
i.e. $\mathtt{\trianglerighteq_{fg}, \rhd_{fg}, \equiv_{fg}}$. To get the
correct behavior we add the constraints which define these relations, see
below. Note, that the relation $\leq$ is purely to prevent duplicates, such as
$\mathtt{f \trianglerighteq g \lor g \trianglerighteq f}$ \textit{and}
$\mathtt{g \trianglerighteq f \lor f \trianglerighteq g}$.

$$\mathtt{\forall_f \forall_{f \leq g} \trianglerighteq_{fg} \lor
\trianglerighteq_{gf}}$$

$$\mathtt{\forall_f \forall_{f \leq g} \equiv_{fg}\ =\ \trianglerighteq_{fg}
\land \trianglerighteq_{gf}}$$

$$\mathtt{\forall_f \trianglerighteq_{ff}}$$

$$\mathtt{\forall_f \forall_{g}\ \rhd_{fg} =\
\trianglerighteq_{fg} \land \neg \trianglerighteq_{gf}}$$

$$\mathtt{\forall_f \forall_{g} \forall_{h}\ \trianglerighteq_{fg} \land
\trianglerighteq_{gh} \to\ \trianglerighteq_{fh}}$$

We define the toprelations $\succsim, \succ, \approx$ on terms by creating a new
variable for every relation and every pair of terms, i.e. $\succsim_{st},
\succ_{st}, \approx_{st}$. To keep everything comprehensible we subdivide the
relations $\succ, \approx$ and create new variables $\succ_{a,st},
\succ_{b,st}, \succ_{c,st}, \approx_{a,st}, \approx_{b,st}$. The toprelations
are now defined using the constraints shown below.

$$\forall_s \forall_t \succsim_{st}\ =\ \succ_{st} \lor \approx_{st}$$
$$\forall_s \forall_t \succ_{st}\ =\ \succ_{a,st} \lor \succ_{b,st} \lor \succ_{c,st}$$
$$\forall_s \forall_t \approx_{st}\ =\ \approx_{a,st} \lor \approx_{b,st}$$

We define constraints for when the types do not match. For example, when
either $s$ or $t$ is a function we restrict $\neg\approx_{a,st}$. We do the
same for all the other subrelations, these constraints follow trivially and
will not be written out. The constraints for when the types do match are
nontrivial. The simple subrelation constraints are as follows.

$$\forall_{s = \mathtt{f}(s_1, \dots, s_n)} \forall_t \succ_{a,st}\ =
\bigvee_{s_i} \succsim_{s_i,t}$$

$$\forall_{s = \mathtt{f}(s_1, \dots, s_n)} \forall_{t = \mathtt{g}(t_1, \dots,
t_m)} \succ_{b,st}\ = \rhd_{\mathtt{f,g}} \land \bigwedge_{t_i} \succ_{s,t_i}$$

$$\forall_s \forall_t \approx_{a,st}\ =\ \approx_{st}$$

We are now left with the two constraints which use the multiset properties. We
will first take a look at $\approx_{b,st}$. Note, when $s$ or $t$ is not a
function, or the number of arguments in $s$ and $t$ are not the same we have
$\neg \approx_{b,st}$. For all of the other cases, where $s=\mathtt{f}(s_1,
\dots, t_n)$ and $t=\mathtt{g}(t_1, \dots, t_n)$, we check whether there exists
a permutation of the arguments of $t$ such that $\forall_n \approx_{s_i,t_i}$.
We define $\mathcal{P}(t_1, \dots, t_n)$ to give all the possible permutations
and use $t^p_i$ to denote $t_i$ in permutation $p \in \mathcal{P}$.

$$\forall_{s = \mathtt{f}(s_1, \dots, s_n)} \forall_{t = \mathtt{g}(t_1, \dots,
t_m)} \approx_{b,st}\ =\ \equiv_{\mathtt{f,g}} \land \bigvee_{p \in
\mathcal{P}(t_1, \dots, t_n)} \bigwedge_i \approx_{s_i,t^p_i}$$

The final constraint $\succ_{c,st}$ is the most elaborate of all. We use the
tip provided in the assignment. We generate all the possible $\mathtt{EQ}$ and
$\varphi$, such that $\exists_{i \in \{1, \dots, n\}}\ \mathtt{EQ}(i) = \bot$
and $\forall_i\ \mathtt{EQ}(\varphi(i)) = \top \to \forall_{j \neq i}\
\varphi(i) \neq \varphi(j)$. We can do this by generating all $\mathtt{EQ}$ and
$\varphi$, then filtering out the functions which do not satisfy the given
constraints. We define $\mathcal{P}^\mathtt{EQ}_\varphi(n, m)$ to give all the
$\mathtt{EQ}, \varphi$ pairs which satisfy the given constraints. The final
constraint is given below, seperated for when $\mathtt{EQ}(\varphi(i)) = \bot$
or $\mathtt{EQ}(\varphi(i)) = \top$.

$$\forall_{s = \mathtt{f}(s_1, \dots, s_n)} \forall_{t = \mathtt{g}(t_1, \dots,
t_m)} \succ_{c,st}\ =\ \equiv_{\mathtt{f,g}} \land \bigvee_{\mathtt{EQ},
\varphi \in \mathcal{P}^\mathtt{EQ}_\varphi(n, m)}
\left(
	\bigwedge_{i,\mathtt{EQ}(\varphi(i)) = \bot} \succ_{s_{\varphi(i)},t_i}
	\bigwedge_{i,\mathtt{EQ}(\varphi(i)) = \top} \approx_{s_{\varphi(i)},t_i}
\right)
$$
\subsection{Results}
Running Z3 with all the constraints and variable definitions as given above
provides us with the following ordering: $\mathtt{a \rhd c \rhd f \rhd g \rhd b
\rhd h \rhd d }$. Finding another symbol ordering is rather easy, we just added
the constraint $\mathtt{\neg \rhd_{h,d}}$. This gives us a new symbol ordering:
$\mathtt{a \rhd c \rhd b \equiv f \rhd g \rhd d \rhd h}$.

The human-readable proof for the inequality
$\mathtt{h}(\mathtt{g}(x,\mathtt{g}(u,z)),\mathtt{c}(x,y,x,z)) \rhd
\mathtt{f}(\mathtt{d}(x,z),u)$ given the first ordering is as follows.
By 2b it suffices to proof that $\mathtt{h \rhd d}$ which holds, and
$\mathtt{h}(\mathtt{g}(x,\mathtt{g}(u,z)),\mathtt{c}(x,y,x,z))~\succ~x$, and
$\mathtt{h}(\mathtt{g}(x,\mathtt{g}(u,z)),\mathtt{c}(x,y,x,z))~\succ~z$. Now
these two inequalities trivially follow from recursively applying rule 2a until
a variable $x$ or $z$ is reached inside the arguments of $\mathtt{h}$, using 3a
on the variables completes the proof.
